---
title: "506 Final Project"
output: html_document
---

*Baggage - The total number of passenger complaints for theft of baggage contents, or for lost, damaged, or misrouted luggage for the airline that month

*Scheduled - The total number of flights scheduled by that airline that month

*Cancelled - The total number of flights cancelled by that airline that month

*Enplaned - The total number of passengers who boarded a plane with the airline that month

```{r}
# load the packages

library(readxl)
library(tidyverse)
library(fpp2)
library(zoo)
library(readr)
library(dplyr)
library(ggplot2)
library(forecast)
set.seed(506)
```


```{r}
df <- read_csv("/Users/UE/Desktop/baggagecomplaints.csv",
                col_types = cols(Date = col_date(format = "%m/%Y")))

df <- df %>%
  mutate(Baggage_Ratio = Baggage / Enplaned *100,
         Cancel_Ratio = Cancelled / Scheduled *100)
df
```

```{r}
bag <- ts(df$Baggage,start = c(2004, 01), end = c(2010, 12), frequency = 12)
autoplot(bag ) +
  theme_classic() +
  labs(x = "Time", y = "Baggage Content Complaints") +
  ggtitle("Baggage Complaint Over Time")
```

```{r}
# Combine the data for different airlines into one time series object
American_Eagle <- subset(df, Airline == 'American Eagle')
Hawaiian <- subset(df, Airline == 'Hawaiian')
United <- subset(df, Airline == 'United')

# Create time series objects for each airline
American_Eagle.ts <- ts(American_Eagle$Baggage, start = c(2004, 1), end = c(2010, 12), frequency = 12)
Hawaiian.ts <- ts(Hawaiian$Baggage, start = c(2004, 1), end = c(2010, 12), frequency = 12)
United.ts <- ts(United$Baggage, start = c(2004, 1), end = c(2010, 12), frequency = 12)

# Merge the time series data into one object
merged_ts <- cbind(American_Eagle = American_Eagle.ts, Hawaiian = Hawaiian.ts, United = United.ts)

# Plot the merged time series data
autoplot(merged_ts) +
  theme_classic() +
  labs(x = "Time", y = "Baggage Complaints") +
  ggtitle("Baggage Complaints Over Time")

```

```{r}
# Calculate baggage complaints per enplaned passenger for each airline
American_Eagle$Baggage_Ratio <- American_Eagle$Baggage / American_Eagle$Enplaned *100
Hawaiian$Baggage_Ratio <- Hawaiian$Baggage / Hawaiian$Enplaned *100
United$Baggage_Ratio <- United$Baggage / United$Enplaned *100

# Combine the ratios into one data frame
merged_ratios <- data.frame(
  American_Eagle = American_Eagle$Baggage_Ratio,
  Hawaiian = Hawaiian$Baggage_Ratio,
  United = United$Baggage_Ratio)

# Convert merged ratios into a time series object
merged_Baggage_Ratio <- ts(merged_ratios, start = c(2004, 1), end = c(2010, 12), frequency = 12)

# Plot the merged ratio time series data
autoplot(merged_Baggage_Ratio) +
  theme_classic() +
  labs(x = "Time", y = "Baggage Complaints per Enplaned Passenger") +
  ggtitle("Baggage Complaints per Enplaned Passenger Over Time")

```


```{R}
# Calculate canceled flights per scheduled flight for each airline
American_Eagle$Cancel_Ratio <- American_Eagle$Cancelled / American_Eagle$Scheduled *100
Hawaiian$Cancel_Ratio <- Hawaiian$Cancelled / Hawaiian$Scheduled *100
United$Cancel_Ratio <- United$Cancelled / United$Scheduled *100

# Combine the ratios into one data frame
merged_Cancel_ratios <- data.frame(
  American_Eagle = American_Eagle$Cancel_Ratio,
  Hawaiian = Hawaiian$Cancel_Ratio,
  United = United$Cancel_Ratio)

# Convert merged ratios into a time series object
merged_Cancel_Ratio <- ts(merged_Cancel_ratios, start = c(2004, 1), end = c(2010, 12), frequency = 12)

# Plot the merged ratio time series data
autoplot(merged_Cancel_Ratio) +
  theme_classic() +
  labs(x = "Time", y = "Canceled Flight divided by Scheduled") +
  ggtitle("Canceled Flights per Scheduled Flight Over Time")
```

```{r}
bag_ratio.ts <- ts(df$Baggage_Ratio,start = c(2004, 01), end = c(2010, 12), frequency = 12)
autoplot(bag_ratio.ts) +
  theme_classic() +
  labs(x = "Time", y = "Baggage per Enplaned Ratio") +
  ggtitle("Baggage per Enplaned Ratio Over Time")
```

```{r}
Cancel_Ratio.ts <- ts(df$Cancel_Ratio,start = c(2004, 01), end = c(2010, 12), frequency = 12)
autoplot(Cancel_Ratio.ts) +
  theme_classic() +
  labs(x = "Time", y = "Flight Cancel Ratio") +
  ggtitle("Flight Cancel Ratioo Over Time")
```



```{r}
stl.bag <- stl(bag, s.window = "periodic")
plot(stl.bag)
```

```{r}
stl.bag_ratio <- stl(bag_ratio.ts, s.window = "periodic")
plot(stl.bag_ratio)
```


```{r}
stl.Cancel_Ratio <- stl(Cancel_Ratio.ts, s.window = "periodic")
plot(stl.Cancel_Ratio)
```

# Data Preproccesing:

```{r}
df <- read_csv("/Users/UE/Desktop/baggagecomplaints.csv",
                col_types = cols(Date = col_date(format = "%m/%Y")))
df
```

```{r}
bag.ts <- ts(df$Baggage,start = c(2004, 01), end = c(2010, 12), frequency = 12)

autoplot(bag.ts) +
  theme_classic() +
  labs(x = "Time", y = "Baggage Complaint Over Time")
```

moving averages

```{r}
# Calculate moving averages
bag.ts.trailing <- rollmean(bag.ts, k = 12, align = "right")
bag.ts.ma.centered <- rollmean(bag.ts, k = 12, fill = NA)  # Using rollmean for centered moving average

# Plotting
plot(bag.ts, ylim = c(1000, 35000), ylab = "Baggage", xlab = "Time", bty = "n", xaxt = "n",
     xlim = c(2004, 2010), main = "")
axis(1, at = seq(2004, 2010, 1), labels = seq(2004, 2010, 1))
lines(bag.ts.ma.centered, lwd = 2, col = "blue")  
lines(bag.ts.trailing, lwd = 2, lty = 2, col = "red") 

# Adding legend
legend("topright", legend = c("Baggage Complaint", "Centered Moving Average", "Trailing Moving Average"),
       lty = c(1, 1, 2), lwd = c(1, 2, 2), col = c("black", "blue", "red"))

```

moving average does not work well with the data that have trend or seasonality. The  moving average does not capture the seasonality in the data. Seasons with high baggage complaints are under-forecasted, and seasons with low baggage complaints are over-forecasted.

simple exponential smoothing model

```{r}
diff.twice.ts <- diff(diff(bag.ts, lag = 12), lag = 4)
training.ts <- window(bag.ts, start = c(2004, 01), end = c(2010, 06))
test.ts <- window(bag.ts, start = c(2010, 07), end = c(2010, 12))

ses <- ets(training.ts, model = "ANN", alpha = 0.2)
ses.pred <- forecast(ses, h = 6, level = 0)
ses.pred

plot(ses.pred, ylim = c(1000, 35000),  ylab = "Baggage Complaints (Twice-Differenced)", xlab = "Time",
    bty = "l", xaxt = "n", xlim = c(2004, 2011), main = "", flty = 2)
axis(1, at = seq(2004, 2011, 1), labels = format(seq(2004, 2011, 1)))
lines(ses.pred$fitted, lwd = 2, col = "blue")
lines(test.ts)

```

```{r}
 ses.opt <- ets(training.ts, model = "ANN")
 ses.opt.pred <- forecast(ses.opt, h = 6, level = 0)
 accuracy(ses.pred, test.ts)
 accuracy(ses.opt.pred, test.ts)
 ses.opt

```

Holt-Winter’s exponential smoothing model 

```{r}
hwin <- ets(training.ts, model = "MAA")
hwin.pred <- forecast(hwin, h = 6, level = 0)

plot(hwin.pred, ylim = c(1000, 35000), ylab = "Baggage Complaints", xlab = "Time",
     bty = "l", xaxt = "n", xlim = c(2004, 2011), main = "", flty = 2)
axis(1, at = seq(2004, 2011, 1), labels = format(seq(2004, 2011, 1)))
lines(hwin.pred$fitted, lwd = 2, col = "blue")
lines(test.ts)
```

```{r}
accuracy(hwin.pred$mean, test.ts)
```




